---
title: "Generative validation for HMM-SSF models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generative validation for HMM-SSF models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

This vignette introduces the minimal workflow implemented in
`HMMSSFGenerativeRepair`. It uses a simulated HMM-SSF-like object so that the
example can run without empirical animal movement data.

The goal is generative validation: simulate trajectories from a fitted model,
then compare the observed trajectory with the simulated trajectories. The main
quantities are diagnostic statistics and Monte Carlo p-values.

## Four trajectory-generation strategies

An HMM-SSF has two coupled layers:

1. a latent state process;
2. a state-specific spatial movement and selection process.

The package separates four ways to generate the latent state path. After the
state path is chosen, spatial steps are generated from the state-specific
movement and selection kernel.

| Method | State path generation | Diagnostic role |
| --- | --- | --- |
| `markov` | Simulate states from the fitted transition model. | Validates the full fitted generative process. |
| `viterbi` | Reuse the Viterbi state path. | Checks whether state-specific kernels work if the decoded path is fixed. |
| `viterbi_tube` | Sample paths close to the Viterbi path under an epsilon tolerance. | Checks sensitivity to near-optimal latent paths. |
| `posterior` | Sample state paths using posterior state information. | Checks sensitivity to latent-state uncertainty. |

The `markov` strategy is the natural full-model generative check. The other
strategies are diagnostic decompositions. They help identify whether a failure
is more likely due to the latent transition model, the decoded state path, or
the state-specific movement kernels.

## Minimal example

```{r setup}
library(HMMSSFGenerativeRepair)

fit <- example_hmmssf_list(n = 50, seed = 123)
observed_track <- fit$observed_track
```

The toy object has two states, a homogeneous transition matrix, and
state-specific movement kernels. It has no habitat-selection covariate; the
movement component is defined directly by the state-specific step-length and
turning-angle distributions. In empirical HMM-SSF applications, the analogous
model summary should report the SSF formula, the transition-probability formula,
the number of available steps, and the state-decoding method.

The call below runs all four simulation strategies and computes a small set of
trajectory-level and state-aware diagnostics.

```{r diagnose}
diagnosis <- diagnose_hmmssf(
  fit = fit,
  observed_track = observed_track,
  n_sims = 9,
  methods = c("markov", "viterbi", "viterbi_tube", "posterior"),
  diagnostics = c(
    "ud_wasserstein",
    "msd",
    "sinuosity",
    "state_occupancy",
    "state_residence_time",
    "state_residence_geometric",
    "switching_rate",
    "transition_counts",
    "state_conditioned_step_length",
    "state_conditioned_turning_angle",
    "state_conditioned_msd"
  ),
  parameter_uncertainty = FALSE,
  epsilon = 10,
  seed = 123
)

head(summary(diagnosis))
```

The dashboard can be drawn for any simulation strategy with the `method`
argument.

```{r dashboard}
plot(diagnosis, method = "markov")
```

The same object contains dashboards for the other strategies:

```{r dashboards-other-methods, eval = FALSE}
plot(diagnosis, method = "viterbi")
plot(diagnosis, method = "viterbi_tube")
plot(diagnosis, method = "posterior")
```

## Reading the Monte Carlo p-values

For each diagnostic, the package compares the observed discrepancy with the
corresponding simulated discrepancies. Small p-values indicate that the
observed trajectory is atypical under the simulated model.

```{r global-pvalues}
global_rows <- is.na(diagnosis$diagnostics$state)
diagnosis$diagnostics[
  global_rows,
  c(
    "method",
    "diagnostic",
    "observed_discrepancy",
    "simulated_discrepancy_median",
    "mc_p_value",
    "mc_p_value_resolution"
  )
]
```

For this example, `n_sims = 9`, so the smallest possible Monte Carlo p-value is
`0.1`. In a general run with `n_sims = N`, the smallest possible p-value is
`1 / (N + 1)`, because the rank p-value uses the finite-simulation correction

```text
(number of simulated statistics at least as extreme as observed + 1) /
  (number of simulations + 1)
```

For diagnostics with missing or non-finite simulated statistics, the effective
resolution is computed from `n_effective`, the number of finite simulated
reference values.

## Reading state diagnostics

State diagnostics are conditional on a state path. In empirical data, this path
is usually decoded, for example with Viterbi, rather than observed directly. The
simulated state paths are generated by the selected simulation strategy.

The diagnostic table therefore includes a `state_reference` column. It is a
reminder that state-level rows compare decoded or supplied observed states with
generated or supplied simulated state paths.

```{r state-reference}
state_rows <- !is.na(diagnosis$diagnostics$state) |
  diagnosis$diagnostics$diagnostic %in% c(
    "state_occupancy_total_variation",
    "switching_rate",
    "transition_counts"
  )

unique(diagnosis$diagnostics[state_rows, "state_reference"])
```

## What the four strategies diagnose

The four strategies should not be interpreted as four competing models. They
are four diagnostic views of the same fitted model.

If `markov` fails but `viterbi` passes, the state-specific movement kernels may
be reasonable, while the fitted transition model may produce unrealistic
state occupancy, switching, or residence times.

If both `markov` and `viterbi` fail, the problem is more likely in the
state-specific movement kernels, omitted covariates, movement distributions, or
state definitions.

If `viterbi_tube` differs from `viterbi`, conclusions are sensitive to small
changes around the decoded latent path.

If `posterior` differs from `viterbi`, conclusions are sensitive to state
classification uncertainty.

## One-state models

For a one-state model, the latent process is degenerate. All four state
generation strategies collapse to the same state path:

```text
1, 1, 1, ..., 1
```

In that case, the meaningful validation is the original trajectory-level
generative validation: compare observed and simulated trajectories using global
diagnostics such as UD, MSD, and sinuosity. State diagnostics such as occupancy,
switching rate, and transition counts are mathematically trivial.

```{r one-state}
one_state_fit <- list(
  initial = 1,
  transition = matrix(1, 1, 1),
  kernels = list(list(
    step = list(dist = "gamma", shape = 2, scale = 0.6),
    angle = list(dist = "wrapped_normal", mean = 0, sd = 0.9)
  )),
  observed_track = observed_track,
  viterbi_path = rep(1L, nrow(observed_track)),
  posterior = matrix(1, nrow = nrow(observed_track), ncol = 1),
  log_emission = matrix(0, nrow = nrow(observed_track), ncol = 1)
)

one_state_diagnosis <- diagnose_hmmssf(
  fit = one_state_fit,
  observed_track = observed_track,
  n_sims = 9,
  methods = c("markov", "viterbi", "viterbi_tube", "posterior"),
  diagnostics = c("ud_wasserstein", "msd", "sinuosity", "state_occupancy"),
  parameter_uncertainty = FALSE,
  epsilon = 0,
  seed = 321
)

one_state_diagnosis$diagnostics[
  one_state_diagnosis$diagnostics$diagnostic %in%
    c("ud_wasserstein", "msd", "sinuosity", "state_occupancy"),
  c("method", "diagnostic", "state", "p_value")
]
```

This is the expected behavior: the trajectory-level diagnostics remain
meaningful, while the state process contributes no additional structure.

## Notes on sharpness

The primary validation quantity is the Monte Carlo p-value. The package also
computes a sharpness label, which summarizes whether the simulated diagnostic
envelope is narrow or diffuse. This label does not replace the p-value. It is a
secondary warning about how informative the simulated reference distribution is.

For workflows that should match a p-value-only validation report, report the
diagnostic statistic, the simulation envelope, and the Monte Carlo p-value.
