---
title: "HSMM-SSF residence-time stress test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HSMM-SSF residence-time stress test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

This vignette studies a deliberately misspecified regime process. The observed
trajectory is generated from an HSMM-like state process with non-geometric dwell
times. The reference model is a best-case homogeneous HMM surrogate: it uses the
same two state labels, the same movement kernels, and a transition matrix
estimated from the generated state path.

This is not a full hidden-state refit. It is a stress test for the residence-time
diagnostics. Because the HMM surrogate is given unusually favorable state
information, a failure of the residence-time diagnostics points directly to the
geometric dwell-time restriction rather than to poor state decoding.

## Simulation design

The example has two states:

| State | Movement interpretation | Dwell-time design |
| --- | --- | --- |
| 1 | Shorter and more tortuous movement | Nearly regular long bouts. |
| 2 | Longer and more directional movement | Nearly regular short bouts. |

There is no habitat-selection covariate in this minimal stress test. The
movement component is a state-specific step-length and turning-angle kernel.
In a fitted HMM-SSF application, this movement component would be replaced by
the fitted SSF formula and covariates.

```{r setup}
library(HMMSSFGenerativeRepair)
library(ggplot2)

wrap_angle_local <- function(x) {
  atan2(sin(x), cos(x))
}

simulate_hsmm_states <- function(n_steps, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }
  states <- integer()
  current <- sample(1:2, size = 1)

  while (length(states) < n_steps) {
    dwell <- if (current == 1L) {
      sample(22:26, size = 1)
    } else {
      sample(7:9, size = 1)
    }
    states <- c(states, rep(current, dwell))
    current <- 3L - current
  }

  states[seq_len(n_steps)]
}

estimate_transition_matrix <- function(states, n_states = 2L, pseudocount = 0.5) {
  counts <- matrix(pseudocount, nrow = n_states, ncol = n_states)
  for (i in seq_len(length(states) - 1L)) {
    counts[states[i], states[i + 1L]] <- counts[states[i], states[i + 1L]] + 1
  }
  counts / rowSums(counts)
}

simulate_track_from_known_states <- function(states, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }
  n_steps <- length(states)
  x <- numeric(n_steps)
  y <- numeric(n_steps)
  bearing <- 0

  for (t in 2:n_steps) {
    state <- states[t]
    step <- if (state == 1L) {
      stats::rgamma(1, shape = 2.4, scale = 0.25)
    } else {
      stats::rgamma(1, shape = 3.0, scale = 0.9)
    }
    turn <- if (state == 1L) {
      stats::rnorm(1, mean = 0, sd = 1.6)
    } else {
      stats::rnorm(1, mean = 0, sd = 0.25)
    }
    bearing <- wrap_angle_local(bearing + turn)
    x[t] <- x[t - 1L] + step * cos(bearing)
    y[t] <- y[t - 1L] + step * sin(bearing)
  }

  data.frame(x = x, y = y)
}
```

## Build the HSMM observation and HMM surrogate

The HMM surrogate is calibrated from the HSMM state path by empirical transition
counts. This gives the HMM the right average switching frequency, but it cannot
reproduce the non-geometric shape of the dwell-time distributions.

```{r simulate-memory-case}
n_steps <- 600
n_sims <- 99

observed_states <- simulate_hsmm_states(n_steps, seed = 100)
observed_track <- simulate_track_from_known_states(observed_states, seed = 101)

hmm_transition <- estimate_transition_matrix(observed_states)
initial <- tabulate(observed_states, nbins = 2)
initial <- initial / sum(initial)

hmm_states <- simulate_markov_states(
  initial = initial,
  transition = hmm_transition,
  n_steps = n_steps,
  n_sims = n_sims,
  seed = 102
)

simulated_tracks <- lapply(seq_len(nrow(hmm_states)), function(i) {
  simulate_track_from_known_states(hmm_states[i, ], seed = 200 + i)
})
names(simulated_tracks) <- paste0("sim_", seq_along(simulated_tracks))

hmm_transition
```

## Diagnose the surrogate HMM

The diagnostic call uses precomputed trajectories and state paths. The
`reference_transition` argument is used only by `state_residence_geometric`.

```{r diagnose-memory}
memory_diagnosis <- diagnose_hmmssf_simulations(
  observed_track = observed_track,
  simulated_tracks = simulated_tracks,
  observed_states = observed_states,
  simulated_states = hmm_states,
  reference_transition = hmm_transition,
  label = "best_case_hmm",
  diagnostics = c(
    "msd",
    "sinuosity",
    "state_occupancy",
    "state_residence_time",
    "state_residence_geometric",
    "switching_rate",
    "transition_counts"
  )
)

selected <- memory_diagnosis$diagnostics[
  memory_diagnosis$diagnostics$diagnostic %in% c(
    "msd",
    "sinuosity",
    "state_occupancy_total_variation",
    "state_residence_time",
    "state_residence_geometric",
    "switching_rate",
    "transition_counts"
  ),
  c(
    "diagnostic",
    "state",
    "observed_discrepancy",
    "simulated_discrepancy_median",
    "mc_p_value",
    "mc_p_value_resolution",
    "interpretation_label"
  )
]

knitr::kable(selected, digits = 3)
```

In this seeded example, occupancy, switching rate, and transition counts are not
extreme because the HMM transition matrix was calibrated from the HSMM state
path. The residence-time diagnostics are extreme because the observed dwell
times are much more regular than the geometric dwell times generated by the HMM.

## Residence-time distributions

```{r residence-distribution-plot}
run_lengths_df <- function(states, source, sim_id = NA_integer_) {
  runs <- rle(states)
  data.frame(
    source = source,
    sim_id = sim_id,
    state = as.integer(runs$values),
    run_length = as.integer(runs$lengths)
  )
}

residence_df <- rbind(
  run_lengths_df(observed_states, "HSMM observation"),
  do.call(
    rbind,
    lapply(seq_len(20), function(i) {
      run_lengths_df(hmm_states[i, ], "HMM simulations", sim_id = i)
    })
  )
)

ggplot(residence_df, aes(x = source, y = run_length, color = source)) +
  geom_jitter(width = 0.15, alpha = 0.45, size = 1.6) +
  facet_wrap(~ state, scales = "free_y") +
  theme_bw() +
  labs(
    x = NULL,
    y = "Residence time",
    color = NULL
  )
```

The plot shows the intended stress-test structure. The HMM can reproduce the
average transition rate, but its residence-time distribution has the geometric
shape implied by the Markov assumption. The HSMM observation has non-geometric
dwell times by construction.

## Interpretation

This vignette supports the following use of the diagnostics:

1. `state_residence_time` is the simulation-based check. It asks whether
   residence times in the decoded or supplied observed state path are unusual
   relative to residence times in trajectories generated by the fitted model.
2. `state_residence_geometric` is the direct homogeneous-HMM check. It asks
   whether decoded residence times are unusual under the geometric dwell-time
   distribution implied by the fitted transition matrix.
3. A small residence-time p-value should be interpreted as evidence of
   residence-time mismatch. It does not by itself identify the correct
   alternative model.

The natural follow-up is to compare the standard HMM-SSF with an HSMM-SSF or
another duration-explicit regime model when residence-time diagnostics fail
while occupancy and transition-count diagnostics remain compatible.
