---
title: "Zebra model comparison with Markov generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zebra model comparison with Markov generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

This article describes a zebra model-comparison workflow. The simulation
strategy is fixed to `markov` for every fitted model. This means that each
candidate model is evaluated through its full generative process:

```text
fitted transition model -> simulated latent states -> state-specific SSF steps
```

Keeping the generation strategy fixed makes the comparison about the fitted
model structure, not about the latent-state conditioning strategy.

The four candidate models are:

| Label | States | Transition model | Interpretation |
| --- | --- | --- | --- |
| `M1_one_state` | 1 | Degenerate | Single-state SSF baseline. |
| `M2_two_state_fixed` | 2 | Constant transition matrix | Two movement states with no time-of-day transition effect. |
| `M3_two_state_time` | 2 | Time-dependent transition matrix | Two states with transition probabilities depending on time of day. |
| `M4_three_state_fixed` | 3 | Constant transition matrix | Three movement states with no time-of-day transition effect. |

The third model is the closest to the zebra HMM-SSF specification used in the
`hmmSSF` zebra example:

```r
tpm_formula <- ~ cos(2 * pi * tod / 24) + sin(2 * pi * tod / 24)
```

The first model is not a genuine HMM transition model. It is the one-state SSF
case: the latent state is always 1, so state diagnostics such as switching rate
and transition counts are mathematically trivial.

## Embedded results

The tables and figures below are the embedded results from
`data-raw/zebra_model_comparison_results.R`. The run uses Markov generation for
all four candidate models.

```{r load-zebra-results}
library(ggplot2)

find_zebra_result <- function(filename) {
  installed_file <- system.file(
    "extdata",
    filename,
    package = "HMMSSFGenerativeRepair"
  )
  if (nzchar(installed_file)) {
    return(installed_file)
  }

  local_candidates <- file.path(
    c("inst/extdata", "../inst/extdata"),
    filename
  )
  local_candidates <- local_candidates[file.exists(local_candidates)]
  if (length(local_candidates) > 0L) {
    return(local_candidates[1L])
  }

  stop("Could not find embedded zebra result file: ", filename, call. = FALSE)
}

model_summary <- utils::read.csv(
  find_zebra_result("zebra_model_comparison_model_summary.csv"),
  stringsAsFactors = FALSE
)
global_diagnostics <- utils::read.csv(
  find_zebra_result("zebra_model_comparison_global_diagnostics.csv"),
  stringsAsFactors = FALSE
)
diagnostics <- utils::read.csv(
  find_zebra_result("zebra_model_comparison_diagnostics.csv"),
  stringsAsFactors = FALSE
)

model_levels <- model_summary$label
plot_model_labels <- c(
  "M1 one-state SSF" = "M1\none-state SSF",
  "M2 two-state fixed TPM" = "M2\n2-state fixed",
  "M3 two-state time TPM" = "M3\n2-state time",
  "M4 three-state fixed TPM" = "M4\n3-state fixed"
)
plot_model_short_labels <- c(
  "M1 one-state SSF" = "M1",
  "M2 two-state fixed TPM" = "M2",
  "M3 two-state time TPM" = "M3",
  "M4 three-state fixed TPM" = "M4"
)
diagnostic_labels <- c(
  ud_wasserstein = "UD Wasserstein",
  msd = "MSD",
  sinuosity = "Sinuosity",
  state_occupancy = "State occupancy",
  state_residence_time = "Residence time",
  state_conditioned_step_length = "Step length by state",
  state_conditioned_turning_angle = "Turning angle by state",
  state_conditioned_msd = "MSD by state"
)

global_diagnostics$model_label <- factor(
  global_diagnostics$model_label,
  levels = model_levels
)
global_diagnostics$diagnostic_label <- factor(
  diagnostic_labels[global_diagnostics$diagnostic],
  levels = diagnostic_labels[c("ud_wasserstein", "msd", "sinuosity")]
)
diagnostics$model_label <- factor(diagnostics$model_label, levels = model_levels)
diagnostics$diagnostic_label <- diagnostic_labels[diagnostics$diagnostic]
```

The simulation settings used for the embedded comparison are:

```{r simulation-settings}
simulation_settings <- unique(model_summary[
  ,
  c("generation", "n_validate", "n_sims", "n_zeros", "rmax")
])

knitr::kable(simulation_settings)
```

The model table reports the fitted model structure and the optimizer status
stored in the fitted objects. In this run, the multi-state `hmmSSF` fits report
`optim` code 1, so these results should be read as a reproducible computational
comparison from these fitted objects, not as final publication-quality maximum
likelihood fits.

```{r fitted-model-summary}
model_summary_display <- model_summary
model_summary_display$convergence_note <- ifelse(
  is.na(model_summary_display$convergence),
  "not applicable",
  ifelse(
    model_summary_display$convergence == 0,
    "converged",
    paste0("optim code ", model_summary_display$convergence)
  )
)

knitr::kable(
  model_summary_display[
    ,
    c("model", "label", "n_states", "tpm", "convergence_note")
  ]
)
```

The next table gives the trajectory-level diagnostics that are comparable across
all four models. A larger Monte Carlo p-value means that the observed statistic
is less extreme relative to the model-generated trajectories.

```{r global-diagnostic-table}
global_display <- global_diagnostics[
  ,
  c(
    "model_label",
    "diagnostic_label",
    "observed_statistic",
    "simulated_median",
    "simulated_q025",
    "simulated_q975",
    "p_value"
  )
]

knitr::kable(
  global_display,
  digits = 3,
  col.names = c(
    "Model",
    "Diagnostic",
    "Observed",
    "Simulated median",
    "Simulated q025",
    "Simulated q975",
    "Monte Carlo p-value"
  )
)
```

```{r global-pvalue-heatmap}
ggplot(
  global_diagnostics,
  aes(x = model_label, y = diagnostic_label, fill = p_value)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", p_value)), size = 3) +
  scale_fill_gradient(
    low = "#D55E00",
    high = "#4C78A8",
    limits = c(0, 1)
  ) +
  scale_x_discrete(labels = plot_model_labels) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8)) +
  labs(
    x = "Model",
    y = "Diagnostic",
    fill = "p-value",
    title = "Zebra observed versus Markov-simulated diagnostics",
    subtitle = "99 Markov simulations per fitted model"
  )
```

```{r observed-simulated-summary}
observed_simulated <- rbind(
  data.frame(
    model_label = global_diagnostics$model_label,
    diagnostic_label = global_diagnostics$diagnostic_label,
    source = "Observed",
    value = global_diagnostics$observed_statistic
  ),
  data.frame(
    model_label = global_diagnostics$model_label,
    diagnostic_label = global_diagnostics$diagnostic_label,
    source = "Simulated median",
    value = global_diagnostics$simulated_median
  )
)

ggplot(
  observed_simulated,
  aes(x = model_label, y = value, color = source, shape = source)
) +
  geom_point(
    size = 2.4,
    position = position_dodge(width = 0.45)
  ) +
  facet_wrap(~ diagnostic_label, scales = "free_y") +
  scale_x_discrete(labels = plot_model_short_labels) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 8),
    plot.margin = margin(10, 20, 20, 20)
  ) +
  labs(
    x = "Model",
    y = "Statistic",
    color = NULL,
    shape = NULL,
    title = "Observed statistic and Markov-simulated median",
    subtitle = "Each panel has its own y-axis scale"
  )
```

The state diagnostics are mainly useful for the multi-state models. For the
one-state baseline, the state process is degenerate and state diagnostics are
shown only as a check that the one-state case is handled.

```{r selected-state-diagnostics}
selected_state_diagnostics <- diagnostics[
  !is.na(diagnostics$state) &
    diagnostics$diagnostic %in% c(
      "state_occupancy",
      "state_residence_time",
      "state_conditioned_step_length",
      "state_conditioned_turning_angle",
      "state_conditioned_msd"
    ),
  ,
  drop = FALSE
]

selected_state_display <- selected_state_diagnostics[
  ,
  c(
    "model_label",
    "diagnostic_label",
    "state",
    "observed_statistic",
    "simulated_median",
    "p_value",
    "warning"
  )
]

knitr::kable(
  selected_state_display,
  digits = 3,
  col.names = c(
    "Model",
    "Diagnostic",
    "State",
    "Observed",
    "Simulated median",
    "Monte Carlo p-value",
    "Warning"
  )
)
```

In this embedded run, no candidate model reproduces every global diagnostic.
The UD Wasserstein p-values are not extreme, but MSD remains extreme for every
model. The three-state fixed model gives the largest p-value for sinuosity in
this run, while the two-state time-dependent model is closest to the original
zebra transition specification. Because the multi-state optimizers report code
1, the comparison should be treated as an executable demonstration and a
diagnostic screen rather than as a final model-selection result.

## Data preparation

The code below assumes that the local `hmmSSF` repository is available and that
the zebra example files can be read from `inst/zebra` and `inst/functions`.
These chunks are not evaluated automatically because fitting the four models and
simulating trajectories is computationally expensive.

```{r zebra-data, eval = FALSE}
library(HMMSSFGenerativeRepair)
library(hmmSSF)
library(raster)
library(ggplot2)
library(survival)

root_dir <- "/Users/aureliennicosia/Documents/validHmmSSF"
hmmssf_repo <- file.path(root_dir, "hmmSSF")
out_dir <- file.path(root_dir, "zebra_validation")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

source(file.path(hmmssf_repo, "inst/functions/cov_df.R"))
source(file.path(hmmssf_repo, "inst/functions/simHMMSSF.R"))

set.seed(250)

zebra_processed <- readRDS(file.path(
  hmmssf_repo,
  "inst/zebra/zebra_processed.RData"
))
track_raw <- subset(zebra_processed, obs == 1)
track_for_controls <- track_raw[, c("ID", "x", "y", "time")]
habitat <- raster::raster(file.path(
  hmmssf_repo,
  "inst/zebra/vegetation2.grd"
))

model_data_file <- file.path(out_dir, "zebra_model_data.rds")

if (file.exists(model_data_file)) {
  model_data <- readRDS(model_data_file)
} else {
  model_data <- hmmSSF::get_controls(
    obs = track_for_controls,
    n_controls = 25,
    distr = "gamma"
  )

  not_na_xy <- which(!is.na(model_data$x) & !is.na(model_data$y))
  model_data$veg[not_na_xy] <- raster::extract(
    habitat,
    model_data[not_na_xy, c("x", "y")],
    method = "simple"
  )
  model_data$veg <- factor(model_data$veg)
  levels(model_data$veg) <- c(
    "grassland",
    "bushed grassland",
    "bushland",
    "woodland"
  )

  model_data$tod <- as.numeric(format(model_data$time, "%H")) +
    as.numeric(format(model_data$time, "%M")) / 60

  zero_step <- which(!is.na(model_data$step) & model_data$step == 0)
  positive_step <- model_data$step[
    !is.na(model_data$step) & model_data$step > 0
  ]
  if (length(zero_step) > 0L) {
    model_data$step[zero_step] <- model_data$step[zero_step] +
      stats::runif(length(zero_step), 0, min(positive_step))
  }

  saveRDS(model_data, model_data_file)
}

observed_full <- subset(model_data, obs == 1)
complete_observed_index <- which(
  is.finite(observed_full$x) & is.finite(observed_full$y)
)

n_validate <- 250
validation_index <- complete_observed_index[seq_len(n_validate)]
observed_validate <- observed_full[
  validation_index,
  c("x", "y", "time", "tod")
]
observed_track <- data.frame(
  x = observed_validate$x,
  y = observed_validate$y
)

ssf_formula <- ~ step + log(step) + cos(angle) + veg
tpm_fixed <- ~ 1
tpm_time <- ~ cos(2 * pi * tod / 24) + sin(2 * pi * tod / 24)
```

## Starting values

The zebra example already contains starting values for the two-state model. For
the fixed-transition models, only transition intercepts are used. For the
three-state model, the third state is initialized as a small perturbation of
the two-state SSF coefficients.

```{r starting-values, eval = FALSE}
initial_par_raw <- readRDS(file.path(
  hmmssf_repo,
  "inst/zebra/initial_par.RData"
))[[3]]

make_three_state_betas <- function(two_state_betas) {
  beta_jitter <- seq(-0.05, 0.05, length.out = nrow(two_state_betas))
  cbind(
    two_state_betas,
    rowMeans(two_state_betas) + beta_jitter
  )
}

make_tpm_start <- function(n_states, tpm_formula, observed_data, raw_alphas) {
  tpm_mm <- stats::model.matrix(tpm_formula, observed_data)
  n_tpm_cov <- ncol(tpm_mm)
  n_cols <- n_states * (n_states - 1L)
  out <- matrix(0, nrow = n_tpm_cov, ncol = n_cols)

  if (n_states == 2L && n_tpm_cov == nrow(raw_alphas)) {
    out[, ] <- raw_alphas
  } else if (n_states == 2L && n_tpm_cov == 1L) {
    out[1L, ] <- raw_alphas[1L, ]
  } else if (n_states == 3L && n_tpm_cov == 1L) {
    out[1L, 1L] <- raw_alphas[1L, 1L]
    out[1L, 3L] <- raw_alphas[1L, 2L]
  }

  out
}
```

## Fitting the four models

`hmmSSF::hmmSSF()` is used for the multi-state HMM-SSF models. The one-state
model is fitted with `survival::clogit()` because it is a standard conditional
logistic SSF with a degenerate latent state.

```{r fit-functions, eval = FALSE}
fit_one_state_ssf <- function(data, ssf_formula) {
  formula <- stats::as.formula(
    paste0("obs ~ ", deparse(ssf_formula[[2L]]), " + strata(stratum)")
  )
  fit <- survival::clogit(formula, data = data)

  betas <- matrix(stats::coef(fit), ncol = 1)
  rownames(betas) <- names(stats::coef(fit))
  colnames(betas) <- "S1"

  list(
    par = list(
      ssf = betas,
      tpm = matrix(numeric(0), nrow = 1, ncol = 0)
    ),
    fit = fit,
    args = list(
      ssf_formula = ssf_formula,
      tpm_formula = ~ 1,
      data = data,
      n_states = 1L
    )
  )
}

fit_hmmssf_model <- function(
  n_states,
  tpm_formula,
  ssf_par0,
  tpm_par0,
  data,
  maxit = 2000
) {
  hmmSSF::hmmSSF(
    ssf_formula = ssf_formula,
    tpm_formula = tpm_formula,
    data = data,
    ssf_par0 = ssf_par0,
    tpm_par0 = tpm_par0,
    n_states = n_states,
    optim_opts = list(trace = 0, maxit = maxit)
  )
}

model_specs <- list(
  M1_one_state = list(
    n_states = 1L,
    tpm_formula = ~ 1,
    fit_type = "one_state_ssf"
  ),
  M2_two_state_fixed = list(
    n_states = 2L,
    tpm_formula = tpm_fixed,
    fit_type = "hmmssf",
    ssf_par0 = initial_par_raw$betas,
    tpm_par0 = make_tpm_start(2, tpm_fixed, observed_validate, initial_par_raw$alphas)
  ),
  M3_two_state_time = list(
    n_states = 2L,
    tpm_formula = tpm_time,
    fit_type = "hmmssf",
    ssf_par0 = initial_par_raw$betas,
    tpm_par0 = make_tpm_start(2, tpm_time, observed_validate, initial_par_raw$alphas)
  ),
  M4_three_state_fixed = list(
    n_states = 3L,
    tpm_formula = tpm_fixed,
    fit_type = "hmmssf",
    ssf_par0 = make_three_state_betas(initial_par_raw$betas),
    tpm_par0 = make_tpm_start(3, tpm_fixed, observed_validate, initial_par_raw$alphas)
  )
)

fits <- lapply(names(model_specs), function(model_name) {
  spec <- model_specs[[model_name]]
  message("Fitting ", model_name)

  if (identical(spec$fit_type, "one_state_ssf")) {
    fit_one_state_ssf(model_data, ssf_formula)
  } else {
    fit_hmmssf_model(
      n_states = spec$n_states,
      tpm_formula = spec$tpm_formula,
      ssf_par0 = spec$ssf_par0,
      tpm_par0 = spec$tpm_par0,
      data = model_data
    )
  }
})
names(fits) <- names(model_specs)
```

## Markov simulation for every model

The simulation strategy is fixed to Markov generation. For the one-state model,
this simply means that every simulated state is 1.

```{r simulate-markov, eval = FALSE}
n_sims <- 99
n_zeros <- 500
rmax <- 5
ssf_cov <- list(veg = habitat)
levels_cov <- list(veg = levels(model_data$veg))
y1 <- as.numeric(observed_validate[1L, c("x", "y")])

simulate_markov_trajectories <- function(model_name, fit, spec) {
  fit_par <- list(
    betas = fit$par$ssf,
    alphas = fit$par$tpm
  )

  tracks_with_states <- lapply(seq_len(n_sims), function(sim_id) {
    simHMMSSF(
      ssf_formula = ssf_formula,
      tpm_formula = spec$tpm_formula,
      ssf_cov = ssf_cov,
      tpm_cov = observed_validate,
      par = fit_par,
      n_states = spec$n_states,
      n_obs = n_validate,
      n_zeros = n_zeros,
      y1 = y1,
      rmax = rmax,
      levels_cov = levels_cov,
      print = FALSE
    )
  })

  simulated_tracks <- lapply(tracks_with_states, function(track) {
    data.frame(x = track$x, y = track$y)
  })
  names(simulated_tracks) <- paste0("sim_", seq_len(n_sims))

  simulated_states <- do.call(
    rbind,
    lapply(tracks_with_states, function(track) as.integer(track$state))
  )

  list(
    simulated_tracks = simulated_tracks,
    simulated_states = simulated_states
  )
}

simulations <- lapply(names(fits), function(model_name) {
  message("Simulating ", model_name)
  simulate_markov_trajectories(
    model_name = model_name,
    fit = fits[[model_name]],
    spec = model_specs[[model_name]]
  )
})
names(simulations) <- names(fits)
```

## Generative diagnostics

The wrapper `diagnose_hmmssf_simulations()` is used here because trajectories
have already been simulated with the zebra-specific simulator.

```{r diagnose-models, eval = FALSE}
diagnostic_names <- c(
  "ud_wasserstein",
  "msd",
  "sinuosity",
  "state_occupancy",
  "state_residence_time",
  "switching_rate",
  "transition_counts",
  "state_conditioned_step_length",
  "state_conditioned_turning_angle",
  "state_conditioned_msd"
)

get_observed_states <- function(model_name, fit, spec) {
  if (spec$n_states == 1L) {
    rep(1L, nrow(observed_track))
  } else {
    hmmSSF::viterbi_decoding(fit)[validation_index]
  }
}

diagnoses <- lapply(names(fits), function(model_name) {
  spec <- model_specs[[model_name]]
  sim <- simulations[[model_name]]

  diagnose_hmmssf_simulations(
    observed_track = observed_track,
    simulated_tracks = sim$simulated_tracks,
    observed_states = get_observed_states(model_name, fits[[model_name]], spec),
    simulated_states = sim$simulated_states,
    label = model_name,
    diagnostics = diagnostic_names,
    metadata = list(
      generation = "markov",
      n_states = spec$n_states,
      tpm_formula = deparse(spec$tpm_formula)
    )
  )
})
names(diagnoses) <- names(fits)

comparison_table <- do.call(
  rbind,
  lapply(names(diagnoses), function(model_name) {
    out <- summary(diagnoses[[model_name]])
    out$model <- model_name
    out
  })
)
```

## Comparing models

The primary comparison is made on the same trajectory-level diagnostics for all
models. State diagnostics are useful for multi-state models, but they are
trivial for the one-state baseline.

```{r comparison-table, eval = FALSE}
global_comparison <- comparison_table[
  is.na(comparison_table$state) &
    comparison_table$diagnostic %in%
      c("ud_wasserstein", "msd", "sinuosity"),
  c("model", "diagnostic", "observed_statistic", "simulated_median", "p_value")
]

global_comparison
```

```{r comparison-plot, eval = FALSE}
ggplot(
  global_comparison,
  aes(x = model, y = diagnostic, fill = p_value)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", p_value)), size = 3) +
  scale_fill_gradient(
    low = "#D55E00",
    high = "#4C78A8",
    limits = c(0, 1)
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    x = "Model",
    y = "Diagnostic",
    fill = "p-value",
    title = "Zebra Markov-generative model comparison"
  )
```

For each model, the full dashboard can also be drawn:

```{r dashboards, eval = FALSE}
plot(diagnoses[["M1_one_state"]], method = "M1_one_state")
plot(diagnoses[["M2_two_state_fixed"]], method = "M2_two_state_fixed")
plot(diagnoses[["M3_two_state_time"]], method = "M3_two_state_time")
plot(diagnoses[["M4_three_state_fixed"]], method = "M4_three_state_fixed")
```

## Interpretation

This workflow answers a targeted question:

```text
If every candidate model is simulated with its own Markov generative process,
which model best reproduces the observed zebra trajectory diagnostics?
```

The comparison is not based on likelihood alone. It asks whether each fitted
model can generate trajectories that reproduce emergent movement features such
as space use, displacement over lag, and path straightness.

For the one-state baseline, only trajectory-level diagnostics should be used
for substantive interpretation. For the multi-state models, state occupancy,
residence time, switching rate, transition counts, and state-conditioned
movement diagnostics help identify why a model succeeds or fails.

The two-state time-dependent model is expected to be the closest to the original
zebra HMM-SSF specification because its transition probabilities vary with time
of day. The fixed-transition models test whether this temporal transition
structure is necessary for generative realism.
